Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-24 related.tex
new draft of related.tex

Modeling time-series data for the purposes of prediction is an old
problem that dates back at least to Yule's 1927 invention of
autoregression~\cite{Yule27}.  Since then hundreds, if not thousands,
of strategies have been developed for a wide variety of prediction
tasks.  The purpose of this paper is not to add a new weapon to this
arsenal, nor to assess or compare the effectiveness of existing
methods.  Our goals are broader: (i) to empirically quantify the
predictive structure that is present in a real-valued scalar time
series and (ii) to explore how that inherent complexity affects the
performance of those existing prediction methods.  It would, of
course, be neither practical nor interesting to report results for
every existing forecast method; instead, we choose a representative
set, which are described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is also not a new problem.  Most solutions fall into
two categories: model-based error analysis and model-free information
analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This can proceed locally or globally.  The local version of
this type of analysis approximates error distributions for different
regions of a time-series model using local ensemble in-sample
forecasting.  {\color{red}[[Joshua, we probably want a quick
      definition of in-sample and out-of-sample.]]} These
distributions are then used to estimate out-of-sample forecast errors
{\color{red}[[in that area of]] [[means?]]} the time series model.
For example, Smith {\sl et al.}  make in-sample forecasts using
ensembles around selected points in the time series in order to
predict the local predictability of a time series reconstructed using
delay-coordinate embedding~\cite{Smith199250}.  This approach can be
used to show that different portions of a time series can exhibit
varying levels of local predictive uncertainty.  We expand on this
finding later in this paper with a time series that exhibits
interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent redundancy of a
time series~\cite{crutchfield2003} and thus cannot be used to draw
conclusions about predictive structure that may be usable by other
forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed {\sl post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to go back
and verify that the model error is normally distributed.  If it is
not, that indicates that there is structure in the time series that
the model-building process was unable to capture and use.
{\color{red} [[The problem with this approach is that if the error is
      normally distributed it can not be said that the practitioner is
      using the ideal forecasting schema---e.g., if nonlinear
      structure is present that linear modeling is oblivious to---but
      only if this class of model used all the structure it was able
      to process.]]  [[having trouble understanding this...]]}

A practice known as deterministic vs. stochastic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global error analysis {\color{red} [[just in
      model-based error analysis?  or also in model-free?]]}  The
basic idea is to construct a series of local linear fits (single
point) to global linear fits (fitting all points known at that time),
increasing the number of points in the fit slowly, and then analyze
how the average out-of-sample error changes as a function of number of
points in the linear fits. The shape of such a graph indicates how
much determinism vs. stochasticity is present in a time series.

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-free information analysis
moves away from those restrictions.  Our approach falls into this
class.  In particular, we empirically measure the inherent complexity
of a time series using an approximation of the Kolmogorov-Sinai
entropy, then correlate that complexity with the predictive accuracy
of forecasts made a number of different schema for that time series.

[[I think we need a short paragraph about redundancy in here somewhere
    --- with a bit more info than was in the intro where we first
    defined it.  We keep referring to it as if the reader knows
    it...]]

Using entropy and redundancy to quantify the inherent predictability
of a time series is not a new idea.  Past methods for this, however,
(e.g.,~\cite{Shannon1951, mantegna1994linguistic}) have hinged on
knowledge of the \emph{generating partition} of the generating
process, which partitions observations into symbols in a way that
preserves the underlying dynamics~\cite{lind95}.  This is a luxury
that is rarely, if ever, afforded to an analyst, since one needs to
know the underlying dynamics in order to construct a generating
partition.  Even if the dynamics are known, these partitions are
difficult to compute and often have fractal
boundaries~\cite{eisele1999}.

Unfortunately, using any partition (such as those produced by simply
binning the data) other than the generation partition induces spurious
complexity to the resulting symbolic sequence, thus misrepresenting
the entropy of the underlying system~\cite{bollt2001}.

It seems desirable to quantify how much information prior observations of a system lend to future values, the \emph{mututal information} between, say, the past $n$ observations at the current one. This quantity is known as the \emph{redundancy}. For example, this would be the amount of information that delay coordinate embedding captures. For IID random processes,  the redundancy is zero, since all observations are independent of each other. On the other hand, any deterministic system --- including chaotic ones --- can be perfectly predicted if observed for long enough~\cite{weigend-book}. Unfortunately it is quite difficult to estimate the redundancy in practice for an arbitrary, real-valued time series. Doing so requires knowing either the Kolmolgorov-Sinai entropy or the \emph{full} positive Lyapunov spectrum. As discussed above, computing the entropy is quite difficult. Similarly, computing the positive Lyapunov exponent(s) from an arbitrary time series is also fraught with difficulties, and is effectively impossible if there is significant noise or the generating system is stochastic.

Our approach expands on \cite{bandt2002per} to approximate normalized redundancy of a time series as a proxy for predictability. This method is different from the current literature because it does not rely on generating partitions, and as such does not suffer from the previously mentioned biases. In addition no knowledge of the underlying generating process is necessary, e.g, linear, nonlinear, the Lyapunov spectrum, etc. In fact, our approach uses no assumption about the underlying generating process, does not require generating partitions, and is applicable to noisy real-valued time series from all classes of systems.

{\color{red}
Maybe talk about ``
wpe has been applied to predicting irregularities in brain wave data but no on one has examined it's correlation with predictive structure. "
but kind of puts the cart before the horse

[[ryan]] i tried writing a paragraph about this around 7 times, and couldn't get anything that sounded reasonable.
}

>>>>

Modeling time-series data for the purposes of prediction dates back at
least to Yule's 1927 invention of autoregression~\cite{Yule27}.  Since
then hundreds, if not thousands, of strategies have been developed for
a wide variety of prediction tasks.  The purpose of this paper is not
to add a new weapon to this arsenal, nor to assess or compare the
effectiveness of existing methods.  Our goals are more general: {\sl
  (i)} to empirically quantify the predictive structure that is
present in a real-valued scalar time series and {\sl (ii)} to explore
how the performance of prediction methods is related to that inherent
complexity.  It would, of course, be neither practical nor interesting
to report results for every existing forecast method; instead, we
choose a representative set, as described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most solutions to it fall
into two categories: model-based error analysis and model-free
information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample forecasting.
{\color{red}[[Joshua, we probably want a quick definition of in-sample
      and out-of-sample.]]} These distributions are then used to
estimate out-of-sample forecast errors {\color{red}[[in that area of]]
  [[means?]]} the time series model.  For example, Smith {\sl et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate the errors and verify that they are normally distributed.
If they are not, that indicates that there is structure in the time
series that the model-building process was unable to capture and use.
The problem with this approach is lack of generality.  Normally
distributed errors indicate that a model has captured the structure in
the data insofar as is possible, \emph{given the formulation of that
  particular model} (viz., the best possible linear fit to a nonlinear
dataset).  This gives no indication as to whether another modeling
strategy might do better.  {\color{red} [[I'm not sure I believe that
      a model's bad fit wouldn't manifest in the error distribution.
      Is that really true?]]}

A practice known as deterministic/sto\-chas\-tic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global model-based error analysis {\color{red} [[or
      also in model-free?]]}  The basic idea is to construct a series
of local linear fits, beginning with a few points and working up to to
a global linear fit that includes all known points, and then analyze
how the average out-of-sample error changes as a function of number of
points in the fit. The shape of such a graph indicates the amounts of
determinism and stochasticity are present in a time series.

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-free information analysis
moves away from those restrictions.  Our approach falls into this
class.  Our goal is to empirically measure the inherent complexity of
a time series, then correlate that complexity with the predictive
accuracy of forecasts made using a number of different methods.

We build on the notion of \emph{redundancy} that was introduced on
page~\pageref{page:redundancy}, which formally quantifies how
information propagates forward through a time series:
% how much information prior observations of a system lend to future
% values
e.g., the mutual information between the past $n$ observations and the
current one.  {\color{red} [[How is redundancy different from/more
      general than MI, then?]]}  For example, this would be the amount
of information that delay coordinate embedding captures.  {\color{red}
  [[This will be very hard to understand, coming as it does before the
      description of that method.  Does it add something essential
      here?]]}  The redundancy of i.i.d. random processes is zero,
since all observations are independent of one another.  On the other
hand, any deterministic system --- including chaotic ones --- can be
perfectly predicted if observed for long enough~\cite{weigend-book}.
{\color{red} [[Previous sentence leaves the connection to redundancy
      implicit.  Need to make it explicit.]]}  
In practice, it is quite difficult to estimate the redundancy of an
arbitrary, real-valued time series.  Doing so requires knowing either
the Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
exponents of the system.  Both of these calculations are
difficult---the latter particularly so if the data are very noisy or
the generating system is stochastic.

Using entropy and redundancy to quantify the inherent predictability
of a time series is not a new idea.  Past methods for this, however,
(e.g.,~\cite{Shannon1951, mantegna1994linguistic}) have hinged on
knowledge of the \emph{generating partition} of the underlying
process, which lets one transform {\color{red}[[what kind of
      observations?  real-valued? scalar?]]}  observations into
symbols in a way that preserves the underlying dynamics~\cite{lind95}.
Different projections---e.g., simply binning the data---can create
spurious complexity in the resulting symbolic sequence and thus
misrepresent the entropy of the underlying system~\cite{bollt2001}.
Generating partitions are luxuries that are rarely, if ever, afforded
to an analyst, since one needs to know the underlying dynamics in
order to construct one.  And even if the dynamics are known, these
partitions are difficult to compute and often have fractal
boundaries~\cite{eisele1999}.

We sidestep these issues by using a variant of the \emph{permutation
  entropy} of Bandt and Pompe~\cite{bandt2002per} to estimate a value
for the Kolmogorov-Sinai entropy of a real-valued time series---and
thus the redundancy in that data, which is an effective proxy for
predictability.  This differs from existing approaches in a number of
ways.  It does not rely on generating partitions---and thus does not
introduce bias into the results if one does not know the dynamics or
cannot compute the partition.  It makes no assumptions about, and
requires no knowledge of, the underlying generating process: linear,
nonlinear, the Lyapunov spectrum, etc.  These features make our
approach applicable to noisy real-valued time series from all classes
of systems.
